A pandemic-scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-09-25 latex/main.tex

Using sc2ts, we built an ARG of $\sim$2.48 million samples spanning the COVID-19 pandemic.
Inference took 21 days on a server equipped with dual EPYC (details) and 500 GB of RAM 
(~5 CPU years; Figure~\ref{fig:inference_resources}).
The ARG contains 2,689,054 nodes representing the genealogical relationships
among 2,482,157 sampled genomes in the Viridian v04 dataset,
which were collected from January 01, 2020 to February 20, 2023 \cite{Hunt2024}.
Of the nodes, 855 represent recombination events,
of which 354 pass quality control (QC) filters
(see STAR Methods, Figure~\ref{fig:recombinant_qc}),
reflecting the overwhelmingly tree-like structure of SARS-CoV-2 evolutionary history.

The ARG represents the full evolutionary history of SARS-CoV-2 and 
is in close agreement with a state-of-the-art UShER phylogeny built using the Viridian dataset \cite{Hunt2024}.
First, we validated the broad structure of the ARG against the UShER phylogeny \cite{Hunt2024}
by pruning both down to representative samples for 1097 deeply-sampled 
lineages (STAR Methods).
Tanglegrams show that the ARG closely matches the phylogenetic
structure of the UShER tree (Figures~\ref{fig:tanglegram_base},
\ref{fig:tanglegram_delta},
\ref{fig:tanglegram_ba2},
\ref{fig:tanglegram_ba5}).
Next, we compared the sc2ts and UShER inferences 
from the Viridian dataset
by pruning both to their intersection of 2,475,418 samples and 27,507 sites,
% JK: this feels a bit wishy washy, but there really is no simple answer to
% this and I don't want to claim we're more parsimonious.
and found they were very similar in terms of parsimony
(Document~\ref{sec:supplement_sc2ts_usher_parsimony})
and imputation of missing and ambiguous bases
(Document~\ref{sec:supplement_sc2ts_usher_imputation}).
We also compared the mutational spectra of major VOCs in the ARG against 
the mutational spectra 
from the Viridian UShER phylogeny \cite{Hunt2024} and a previous study \cite{Bloom2023} (STAR Methods),
finding a close match to them (Figure~\ref{fig:mutational_spectra}).

Input genome alignments are masked in sc2ts
so that deletions and ambiguous bases are treated as missing data by the HMM,
and once the samples are incorporated into the ARG,
these bases are imputed to carry the values inherited by the node in question.
Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023}, however,
and to incorporate them into the ARG we identified a set of [XXX] sites
most frequently involved in deletions [cite source?]
and post-hoc mapped alignment data for these sites to the ARG using parsimony.
Although sites are treated independently in this approach,
the approximation successfully identified the most frequent deletions,
such as the recurrent 21991-21993 (See Document~\ref{sec:supplement_dels}
for additional analysis).
The node representing the origin of Alpha, with 290,407 descendants,
is correctly associated with key lineage-defining deletions:
11288-11296 (ORF1A del3675-3677),
21765-21770 (Spike del69-70),
21991-21993 (Spike del144), and
28271 (non-coding).
The Spike 69-70 deletion, which causes S-Gene Target Failure \cite{Walker2021},
also emerges twice in the origin of Omicron -
in the node leading to the BA.1 variant (342,099 descendants),
and the node leading to the BA.4 variant (189,746 descendants).
To speed up inference and to avoid issues caused by sites enriched for errors [citation?],
sc2ts masks out a set of 100 sites identified as the most highly homoplasic
by running on a subset of the data (STAR Methods).
Of these sites, 9 are in the set of [1XX] ``problematic sites'' 
masked by UShER \cite{Hunt2024}.
In addition to remapping the data for sites identified as being involved in deletions,
we added mutations for XX of these highly recurrent sites
that are used in Pango designation \cite{Rambaut2020} by post-hoc parsimony.
Each node in the ARG was then assigned to one of 2058 Pango lineages using pangolin (STAR Methods).
All but 21,680 sampled sequences (0.87\%) agreed with the source sequences in terms of Pango lineage assignment
(comparable to the 0.27\% disagreement between pangolin-data 
versions 1.21 and 1.29 in the source Viridian metadata).
A majority of these lineages (1473 of 2058) have monophyletic origins within the ARG,
increasing to 1779 (86\%) when we allow for multiple sibling origination 
nodes (Document~\ref{sec:supplement_pango_lineages}).

>>>>

Using sc2ts, we built an ARG from 2,482,157 samples from the Viridian v04
dataset\cite{Hunt2024}, collected between 2020-01-01 and 2023-02-20.
Inference took 21 days on a XX core server with 500 GB of RAM 
($\sim 5$ CPU years; Figure~\ref{fig:inference_resources}).
After postprocessing, the ARG 
contains 2,689,054 nodes with estimated dates, 
2,748,838 edges, and 2,285,344 mutations at 29,893 sites. 
Including sample IDs and the Pango and Scorpio designations computed by
Pangolin~\cite{OToole2021assignment} as metadata for each node,
the ARG is stored in a 32MiB tszip format file.
Of the nodes, 855 represent recombination events, 
resulting in 316 distinct trees along the genome (some
events occured at the same breakpoints).
% It's confusing to talk about recombinant QC here as we're talking 
% about the full ARG for the rest of the section.

The ARG is a comprehesive record of the evolutionary history of SARS-CoV-2 and 
is in close agreement with a state-of-the-art UShER phylogeny also built 
from the Viridian dataset \cite{Hunt2024}.
First, we validated the broad structure of the ARG against 
the UShER phylogeny \cite{Hunt2024}
by pruning both down to representative samples for 1097 deeply-sampled 
lineages (STAR Methods).
Tanglegrams show that the ARG closely matches the phylogenetic
structure of the UShER tree (Figures~\ref{fig:tanglegram_base},
\ref{fig:tanglegram_delta},
\ref{fig:tanglegram_ba2},
\ref{fig:tanglegram_ba5}).
Next, we compared the sc2ts and UShER inferences 
from the Viridian dataset
by pruning both to their intersection of 2,475,418 samples and 27,507 sites,
% JK: this feels a bit wishy washy, but there really is no simple answer to
% this and I don't want to claim we're more parsimonious.
and found they were very similar in terms of parsimony
(Document~\ref{sec:supplement_sc2ts_usher_parsimony})
and imputation of missing and ambiguous bases
(Document~\ref{sec:supplement_sc2ts_usher_imputation}).
We also compared the mutational spectra of major VOCs in the ARG against 
the mutational spectra 
from the Viridian UShER phylogeny and a previous study \cite{Bloom2023} (STAR Methods),
finding close agreement (Figure~\ref{fig:mutational_spectra}).

% FIXME this is too detailed, and too methodsy. Abstract out and summarise
% results
Input genome alignments are masked in sc2ts
so that deletions and ambiguous bases are treated as missing data by the HMM,
and once the samples are incorporated into the ARG,
these bases are imputed to carry the values inherited by the node in question.
Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023}, however,
and to incorporate them into the ARG we identified a set of [XXX] sites
most frequently involved in deletions [cite source?]
and post-hoc mapped alignment data for these sites to the ARG using parsimony.
Although sites are treated independently in this approach,
the approximation successfully identified the most frequent deletions,
such as the recurrent 21991-21993 (See Document~\ref{sec:supplement_dels}
for additional analysis).
The node representing the origin of Alpha, with 290,407 descendants,
is correctly associated with key lineage-defining deletions:
11288-11296 (ORF1A del3675-3677),
21765-21770 (Spike del69-70),
21991-21993 (Spike del144), and
28271 (non-coding).
The Spike 69-70 deletion, which causes S-Gene Target Failure \cite{Walker2021},
also emerges twice in the origin of Omicron -
in the node leading to the BA.1 variant (342,099 descendants),
and the node leading to the BA.4 variant (189,746 descendants).
To speed up inference and to avoid issues caused by sites enriched for errors [citation?],
sc2ts masks out a set of 100 sites identified as the most highly homoplasic
by running on a subset of the data (STAR Methods).
Of these sites, 9 are in the set of [1XX] ``problematic sites'' 
masked by UShER \cite{Hunt2024}.
In addition to remapping the data for sites identified as being involved in deletions,
we added mutations for XX of these highly recurrent sites
that are used in Pango designation \cite{Rambaut2020} by post-hoc parsimony.
Each node in the ARG was then assigned to one of 2058 Pango lineages using pangolin (STAR Methods).
All but 21,680 sampled sequences (0.87\%) agreed with the source sequences in terms of Pango lineage assignment
(comparable to the 0.27\% disagreement between pangolin-data 
versions 1.21 and 1.29 in the source Viridian metadata).
A majority of these lineages (1473 of 2058) have monophyletic origins within the ARG,
increasing to 1779 (86\%) when we allow for multiple sibling origination 
nodes (Document~\ref{sec:supplement_pango_lineages}).
